---
title: "BLUEs_System_GxE"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r, message=FALSE}
library(readxl) #for loading data from excel files
library(tidyverse) #for data manipulation
library(lme4) #for linear models and BLUEs calculation
```

# Load data

Here we load the required data from the raw_data folder within this
project.

```{r}
#load excel file of phenotype data
phenotype <- read.csv(
  file = "raw_data/phenotype/Alldata2016-2018 3sd NA_FINAL_NO_ISSUES.csv") 
```

# Add index column

For subsequent analysis it will be helpful to add an index for the varieties to
save typing out full variety names.

```{r}
phenotype <- phenotype %>%
  mutate(Genotype_ID = rep(1:282, each = 12)) %>% #282 vars each repeated x12
  relocate(Genotype_ID, before = Variety.name) %>% #move before variety.name col
  rename(Variety.name = before) #rename before col back to variety.name
         
head(phenotype[1:6]) #check result
```

# Traits of interest

Since we have data for 140 different traits we will focus on those of particular
interest.

These traits are:

- Yield (Kg)
- Max Max height (m) = maximum recorded height for max/highest point of plant
- Max Ave height (m) = maximum average height
- Max_area_plant (m2) = maximum plant area
- Max_volume_plant (m3) = max plant volume

First create a vector of traits of interest for future use.

```{r}
traits_of_interest <- phenotype %>%
  select(Yield_Kg, 
         Max_Max_height_m,
         Max_Ave_height_m,
         Max_area_plant_m2,
         Max_volume_plant_m3) %>%
  colnames()
```

Since certain traits in the data are not measured every year we need to check 
when our traits of interest have been measured. We can do this as follows...

### 2016

```{r}
phenotype %>%
  filter(Year == 2016) %>% #filter by year
  select(all_of(traits_of_interest)) %>% #select the traits of interest
  sapply(., \(x) is.na(x)) %>% #logical vector checking if entries are NA
  as.data.frame() %>% #convert to df
  colSums() %>% #calculate colsums i.e. if no NAs will equal 0
  as.data.frame() %>% #convert back to df
  filter(. < 1128) #>1128 NAs = the year is missing as 1128 rows per year
```
Based on this only Yield_Kg has the data required for analysis in 2016.

### 2017
```{r}
phenotype %>%
  filter(Year == 2017) %>% #filter by year
  select(all_of(traits_of_interest)) %>% #select the traits of interest
  sapply(., \(x) is.na(x)) %>% #logical vector checking if entries are NA
  as.data.frame() %>% #convert to df
  colSums() %>% #calculate colsums i.e. if no NAs will equal 0
  as.data.frame() %>% #convert back to df
  filter(. < 1128) #>1128 NAs = the year is missing as 1128 rows per year
```

Based on this all traits of interest are measured in 2017.

### 2018

```{r}
phenotype %>%
  filter(Year == 2018) %>% #filter by year
  select(all_of(traits_of_interest)) %>% #select the traits of interest
  sapply(., \(x) is.na(x)) %>% #logical vector checking if entries are NA
  as.data.frame() %>% #convert to df
  colSums() %>% #calculate colsums i.e. if no NAs will equal 0
  as.data.frame() %>% #convert back to df
  filter(. < 1128) #>1128 NAs = the year is missing as 1128 rows per year
```

Based on this all traits of interest are measured in 2018.

Overall yield is measured across all three years, the other traits of interest
are only measured in 2017 and 2018.

We will create a subset of traits in 2017 and 2018 for subsequent analysis.

```{r}
phenotype_17_18 <- phenotype %>%
  filter(Year == 2017 | Year == 2018)
```

# Replace NA values

As demonstrated for the above, even when the traits are measured for a given
year there are some values missing. Thus, we now need to check for and 
replace NAs for each of the different traits.

### Check for and replace NAs for yield_kg

Unlike the other traits, yield is measured across all three years so we 
will just use the full dataset.

```{r}
phenotype %>%
  filter(is.na(Yield_Kg)) %>% #select NA values
  select(Genotype_ID) %>% #display genotype ID col
  pull() #show as vector
```

Based on this genotypes/varieties 93 and 115 are NA. We can see this in more
detail below...

```{r}
phenotype %>%
  rownames_to_column() %>%
  filter(Genotype_ID == 93 
         | Genotype_ID == 115) %>% #filter to genotypes with NA for yield
  select(rowname,
         System,
         Genotype_ID,
         Variety.name, 
         Replicate, 
         Year, 
         Yield_Kg) #select info cols + yield trait
```

Looking at this the NAs for are in rows 1114 (ID 93) and 1378 (ID 115). 
They are also both for Organic, Replicate 2 in 2018. Thus to account for this 
missing data we will just replace the NAs with Organic, Replicate 1 in 2018 for 
each variety.

```{r}
#select replacement data for genotype 93
yield_ID93 <- phenotype %>%
  filter(Genotype_ID == 93 
         & System == "Org" 
         & Replicate == 1
         & Year == 2018) %>%
  select(Yield_Kg) %>%
  pull()

#replace NA for genotype 93
phenotype$Yield_Kg[1114] <- yield_ID93

#select replacement data for genotype 115
yield_ID115 <- phenotype %>%
  filter(Genotype_ID == 115 
         & System == "Org" 
         & Replicate == 1
         & Year == 2018) %>%
  select(Yield_Kg) %>%
  pull()

#replace NA for genotype 115
phenotype$Yield_Kg[1378] <- yield_ID115
```

### Check for and replace NAs for height, area, volume traits.

Only measured in 2017 and 18 so use the subset of data. Done all together as 
same number of NAs for each trait = likely the same values missing.

```{r}
#Max_Max_height_m
phenotype_17_18 %>%
  filter(is.na(Max_Max_height_m)) %>% #select NA values
  select(Genotype_ID) %>% #display genotype ID col
  rename(Max_Max_height_m_NAs = Genotype_ID) #rename col to make output clear

#Max_Ave_height_m
phenotype_17_18 %>%
  filter(is.na(Max_Ave_height_m)) %>% #select NA values
  select(Genotype_ID) %>% #display genotype ID col
  rename(Max_Ave_height_m_NAs = Genotype_ID) #rename col to make output clear

#Max_area_plant_m2
phenotype_17_18 %>%
  filter(is.na(Max_area_plant_m2)) %>% #select NA values
  select(Genotype_ID) %>% #display genotype ID col
  rename(Max_area_plant_m2_NAs = Genotype_ID) #rename col to make output clear

#Max_volume_plant_m3
phenotype_17_18 %>%
  filter(is.na(Max_volume_plant_m3)) %>% #select NA values
  select(Genotype_ID) %>% #display genotype ID col
  rename(Max_volume_plant_m3_NAs = Genotype_ID) #rename col to make output clear
```

In the case of all these traits genotype 93 has one NA, 148 has two. 
We can see this in more detail below...

```{r}
phenotype_17_18 %>%
  rownames_to_column() %>%
  filter(Genotype_ID == 93 
         | Genotype_ID == 148) %>% #filter to genotypes with NA 
  filter(is.na(Max_Max_height_m)) %>% #filter to just show NAs
  select(rowname,
         System,
         Genotype_ID,
         Variety.name, 
         Replicate, 
         Year, 
         Max_Max_height_m,
         Max_Ave_height_m,
         Max_area_plant_m2,
         Max_volume_plant_m3) #select info cols + traits
```

Thus, for all these traits...

The NA for ID 93 is in row 742 for Organic, Replicate 2 in 2018. 
We can just replace these NA with Organic, Replicate 1 in 2018 for this variety.

```{r}
#Max_Max_height_m

#select replacement data for genotype 93
Max_Max_height_m_ID93 <- phenotype_17_18 %>%
  filter(Genotype_ID == 93 
         & System == "Org" 
         & Replicate == 1
         & Year == 2018) %>%
  select(Max_Max_height_m) %>%
  pull()

#replace NA for genotype 93
phenotype_17_18$Max_Max_height_m[742] <- Max_Max_height_m_ID93

#Max_Ave_height_m

#select replacement data for genotype 93
Max_Ave_height_m_ID93 <- phenotype_17_18 %>%
  filter(Genotype_ID == 93 
         & System == "Org" 
         & Replicate == 1
         & Year == 2018) %>%
  select(Max_Ave_height_m) %>%
  pull()

#replace NA for genotype 93
phenotype_17_18$Max_Ave_height_m[742] <- Max_Ave_height_m_ID93

#Max_area_plant_m2

#select replacement data for genotype 93
Max_area_plant_m2_ID93 <- phenotype_17_18 %>%
  filter(Genotype_ID == 93 
         & System == "Org" 
         & Replicate == 1
         & Year == 2018) %>%
  select(Max_area_plant_m2) %>%
  pull()

#replace NA for genotype 93
phenotype_17_18$Max_area_plant_m2[742] <- Max_area_plant_m2_ID93

#Max_volume_plant_m3

#select replacement data for genotype 93
Max_volume_plant_m3_ID93 <- phenotype_17_18 %>%
  filter(Genotype_ID == 93 
         & System == "Org" 
         & Replicate == 1
         & Year == 2018) %>%
  select(Max_volume_plant_m3) %>%
  pull()

#replace NA for genotype 93
phenotype_17_18$Max_volume_plant_m3[742] <- Max_volume_plant_m3_ID93
```

The NAs for ID 148 are both in Organic 2017 (rows 1177 and 1178). Since there is 
no replicate will replace both these NAs with mean for Organic 2018.

```{r}
#Max_Max_height_m

#select replacement data for ID 148
Max_Max_height_m_ID148 <- phenotype_17_18 %>%
  filter(Genotype_ID == 148 
         & System == "Org" 
         & Year == 2018) %>%
  select(Max_Max_height_m) %>%
  colMeans()

#replace NA for genotype 148
phenotype_17_18$Max_Max_height_m[1177] <- Max_Max_height_m_ID148
phenotype_17_18$Max_Max_height_m[1178] <- Max_Max_height_m_ID148

#Max_Ave_height_m

#select replacement data for ID 148
Max_Ave_height_m_ID148 <- phenotype_17_18 %>%
  filter(Genotype_ID == 148 
         & System == "Org" 
         & Year == 2018) %>%
  select(Max_Ave_height_m) %>%
  colMeans()

#replace NA for genotype 148
phenotype_17_18$Max_Ave_height_m[1177] <- Max_Ave_height_m_ID148
phenotype_17_18$Max_Ave_height_m[1178] <- Max_Ave_height_m_ID148

#Max_area_plant_m2

#select replacement data for ID 148
Max_area_plant_m2_ID148 <- phenotype_17_18 %>%
  filter(Genotype_ID == 148 
         & System == "Org" 
         & Year == 2018) %>%
  select(Max_area_plant_m2) %>%
  colMeans()

#replace NA for genotype 148
phenotype_17_18$Max_area_plant_m2[1177] <- Max_area_plant_m2_ID148
phenotype_17_18$Max_area_plant_m2[1178] <- Max_area_plant_m2_ID148

#Max_volume_plant_m3

#select replacement data for ID 148
Max_volume_plant_m3_ID148 <- phenotype_17_18 %>%
  filter(Genotype_ID == 148 
         & System == "Org" 
         & Year == 2018) %>%
  select(Max_volume_plant_m3) %>%
  colMeans()

#replace NA for genotype 148
phenotype_17_18$Max_volume_plant_m3[1177] <- Max_volume_plant_m3_ID148
phenotype_17_18$Max_volume_plant_m3[1178] <- Max_volume_plant_m3_ID148
```

# BLUEs

Having selected the traits of interest the next step is to calculate
Best Linear Unbiased estimates for each trait.

For the original BLUEs (i.e. not System_GxE) this is done by setting up a mixed effect model where...

- Fixed effects = Trait of interest, Genotype (i.e. Trait ~ Genotype)
- Random effects = Cultivation system (System) and Year

For the System_GxE BLUEs we shall also add 

- Gene x System as a random effect (Genotype:System)

Note that in order for the model to work in lme4 (the package used modelling) 
we must ensure System, Year and Genotype_ID are set to factors. 

For full data-set (when doing yield_kg trait)

```{r}
phenotype$Genotype_ID <- as.factor(phenotype$Genotype_ID)
phenotype$System <- as.factor(phenotype$System)
phenotype$Year <- as.factor(phenotype$Year)
```

For 2017/18 data-set (when doing other traits)

```{r}
phenotype_17_18$Genotype_ID <- as.factor(phenotype_17_18$Genotype_ID)
phenotype_17_18$System <- as.factor(phenotype_17_18$System)
phenotype_17_18$Year <- as.factor(phenotype_17_18$Year)
```

# Yield (Kg)

Yield is measured across all three years so we can just use the full dataset.

### Calculate BLUEs

```{r}
#note here have used REML = TRUE which gives less biased variance estimates
#for random effects which is better for calculating BLUEs

yield_lmer <- lmer(formula = Yield_Kg ~ Genotype_ID + (1|System) + (1|Year) + (1|Genotype_ID:System),
                   data = phenotype,
                   REML = TRUE)
```

```{r}
#note fixef(yield_lmer) is essentially the same as yield_lmer@beta it just gives
#a named vector instead (fixef(yield_lmer) = ID/name + fixed effect estimate, 
#yield_lmer@beta = fixed effect estimate).

fe <- fixef(yield_lmer) #extract fixed effects coefficients/estimates

BLUEs <- fe+c(0,rep(fe[1],length(fe)-1)) #add intercept to all subsequent values

yield_BLUEs <- 
  BLUEs %>% #BLUEs data as vector
  as_tibble() %>% #convert to tibble
  mutate(ID = unique(phenotype$Variety.name)) %>% #add variety name col
  relocate(ID, before = value) %>% #move ID col to front (for GWAS_poly)
  rename(yield_kg = before) #rename col to yield_kg
```

### QQ plots

```{r, message=FALSE}
pdf(file = "System_GxE_BLUEs_output/yield_qq.pdf")

residuals(yield_lmer) %>%
  as_tibble %>%
  ggplot(mapping = aes(x = value))+
  geom_histogram(binwidth = 0.4)

qqnorm(residuals(yield_lmer))

qqline(residuals(yield_lmer))

plot(yield_lmer)

dev.off()
```

# Max_Max_height_m

### Calculate BLUEs

```{r}
Max_Max_height_m_lmer <- lmer(formula = Max_Max_height_m ~ Genotype_ID 
                         + (1|System) + (1|Year) + (1|Genotype_ID:System),
                         data = phenotype_17_18,
                         REML = TRUE)
```

```{r}
fe <- fixef(Max_Max_height_m_lmer) #extract fixed effects estimates

BLUEs <- fe+c(0,rep(fe[1],length(fe)-1)) #add intercept to all subsequent values

Max_Max_height_m_BLUEs <- 
  BLUEs %>% #BLUEs data as vector
  as_tibble() %>% #convert to tibble
  mutate(ID = unique(phenotype$Variety.name)) %>% #add variety name col
  relocate(ID, before = value) %>% #move ID col to front (for GWAS_poly)
  rename(Max_Max_height_m = before) #rename col to Max_Max_height_m
```

### QQ plots

```{r, message=FALSE}
pdf(file = "System_GxE_BLUEs_output/Max_Max_height_m_qq.pdf")

qqnorm(residuals(Max_Max_height_m_lmer))

qqline(residuals(Max_Max_height_m_lmer))

plot(Max_Max_height_m_lmer)

dev.off()
```

# Max_Ave_height_m

### Calculate BLUEs

```{r}
Max_Ave_height_m_lmer <- lmer(formula = Max_Ave_height_m ~ Genotype_ID 
                              + (1|System) + (1|Year) + (1|Genotype_ID:System),
                              data = phenotype_17_18,
                              REML = TRUE)
```

```{r}
fe <- fixef(Max_Ave_height_m_lmer) #extract fixed effects estimates

BLUEs <- fe+c(0,rep(fe[1],length(fe)-1)) #add intercept to all subsequent values

Max_Ave_height_m_BLUEs <- 
  BLUEs %>% #BLUEs data as vector
  as_tibble() %>% #convert to tibble
  mutate(ID = unique(phenotype$Variety.name)) %>% #add variety name col
  relocate(ID, before = value) %>% #move ID col to front (for GWAS_poly)
  rename(Max_Ave_height_m = before) #rename col to Max_Ave_height_m
```

### QQ plots

```{r, message=FALSE}
pdf(file = "System_GxE_BLUEs_output/Max_Ave_height_m_qq.pdf")

qqnorm(residuals(Max_Ave_height_m_lmer))

qqline(residuals(Max_Ave_height_m_lmer))

plot(Max_Ave_height_m_lmer)

dev.off()
```

# Max_area_plant_m2

### Calculate BLUEs

```{r}
Max_area_plant_m2_lmer <- lmer(formula = Max_area_plant_m2 ~ Genotype_ID 
                              + (1|System) + (1|Year) + (1|Genotype_ID:System),
                              data = phenotype_17_18,
                              REML = TRUE)
```

```{r}
fe <- fixef(Max_area_plant_m2_lmer) #extract fixed effects estimates

BLUEs <- fe+c(0,rep(fe[1],length(fe)-1)) #add intercept to all subsequent values

Max_area_plant_m2_BLUEs <- 
  BLUEs %>% #BLUEs data as vector
  as_tibble() %>% #convert to tibble
  mutate(ID = unique(phenotype$Variety.name)) %>% #add variety name col
  relocate(ID, before = value) %>% #move ID col to front (for GWAS_poly)
  rename(Max_area_plant_m2 = before) #rename col to Max_area_plant_m2
```

### QQ plots

```{r, message=FALSE}
pdf(file = "System_GxE_BLUEs_output/Max_area_plant_m2_qq.pdf")

qqnorm(residuals(Max_area_plant_m2_lmer))

qqline(residuals(Max_area_plant_m2_lmer))

plot(Max_area_plant_m2_lmer)

dev.off()
```

# Max_volume_plant_m3

### Calculate BLUEs

```{r}
Max_volume_plant_m3_lmer <- lmer(formula = Max_volume_plant_m3 ~ Genotype_ID 
                               + (1|System) + (1|Year) + (1|Genotype_ID:System),
                               data = phenotype_17_18,
                               REML = TRUE)
```

```{r}
fe <- fixef(Max_volume_plant_m3_lmer) #extract fixed effects estimates

BLUEs <- fe+c(0,rep(fe[1],length(fe)-1)) #add intercept to all subsequent values

Max_volume_plant_m3_BLUEs <- 
  BLUEs %>% #BLUEs data as vector
  as_tibble() %>% #convert to tibble
  mutate(ID = unique(phenotype$Variety.name)) %>% #add variety name col
  relocate(ID, before = value) %>% #move ID col to front (for GWAS_poly)
  rename(Max_volume_plant_m3 = before) #rename col to Max_volume_plant_m3
```

### QQ plots

```{r, message=FALSE}
pdf(file = "System_GxE_BLUEs_output/Max_volume_plant_m3_qq.pdf")

qqnorm(residuals(Max_volume_plant_m3_lmer))

qqline(residuals(Max_volume_plant_m3_lmer))

plot(Max_volume_plant_m3_lmer)

dev.off()
```

# Save BLUES

```{r}

#merge dfs
all_BLUEs <- cbind(yield_BLUEs,
      Max_Max_height_m_BLUEs,
      Max_Ave_height_m_BLUEs,
      Max_area_plant_m2_BLUEs,
      Max_volume_plant_m3_BLUEs)

#remove additional ID cols
all_BLUEs <- all_BLUEs[-c(3,5,7,9)]

write.csv(all_BLUEs, 
          "System_GxE_BLUEs_output/System_GxE_BLUEs.csv",
          row.names = FALSE)
```

Also save BLUEs with Q-matrices from STRUCTURE and DAPC
in a single file.

```{r}
#load and prep DAPC membership probabilities/Q-matrix
DAPC_Q_matrix <- read.csv(file = "DAPC_output/DAPC_membership_probabilities.csv") %>%
  rename(DAPC_cluster_1 = X1) %>%
  rename(DAPC_cluster_2 = X2) %>%
  rename(DAPC_cluster_3 = X3) %>%
  rename(DAPC_cluster_4 = X4)

#merge BLUEs and DAPC membership probabilities/Q-matrix
BLUEs_with_DAPC_Q_matrix <- cbind(all_BLUEs, DAPC_Q_matrix)
BLUEs_with_DAPC_Q_matrix <- 
  BLUEs_with_DAPC_Q_matrix[-c(7)] #remove additional ID col

################

#load STRUCTURE Q-matrix
STRUCTURE_Q_matrix <- read.table("STRUCTURE/final_results/bestK/CLUMPP/K3.outfile")

#prep STRUCTURE Q-matrix
STRUCTURE_Q_matrix <- STRUCTURE_Q_matrix[6:8] %>%
  rename("STRUCTURE_K1" = V6) %>%
  rename("STRUCTURE_K2" = V7) %>%
  rename("STRUCTURE_K3" = V8)

#merge BLUEs with STRUCTURE Q-matrix
BLUEs_with_Q_matrices <- cbind(BLUEs_with_DAPC_Q_matrix, STRUCTURE_Q_matrix) 

################

#load and prep PCs from PCA
PCs <- read.csv(file = "DAPC_output/PCA_principal_components.csv") %>%
  select(-c("ID")) #remove ID col as not needed 

#merge BLUEs with PCs
BLUEs_with_Q_matrices_and_PCs <- cbind(BLUEs_with_Q_matrices, PCs)

#save final file
write.csv(BLUEs_with_Q_matrices_and_PCs, 
          "System_GxE_BLUEs_output/System_GxE_BLUEs_with_Q_matrices_and_PCs.csv",
          row.names = FALSE)

```

# References

BATES, D., MÄCHLER, M., BOLKER, B., and WALKER, S. (2015). ‘Fitting Linear Mixed-Effects Models Using lme4,’ Journal of Statistical Software, 67(1), pp. 1 - 48. doi:10.18637/jss.v067.i01.

YANG, C.J., RUSSELL, J., RAMSAY, L., THOMAS, W., POWELL, W., and MACKAY, I. (2021). ‘Overcoming barriers to the registration of new plant varieties under the DUS system,’ Communications Biology, 4(1), pp. 302. doi:10.1038/s42003-021-01840-9.

https://github.com/cjyang-work/DUS = github for scripts in yang et al., 2021



